Geometric models reveal behavioral and neural signatures of transforming experiences into memories

### 2020-10-14 paper/main.tex
added confidence intervals to in-text statistics

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both Chen et al. (2017)~\cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity~\citep{MannKaha12}.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.005$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015$) numbers of recalled events.

@@

In addition to examining how precisely and distinctively participants recalled individual events, one may also use these metrics to summarize each participant's performance by averaging across a participant's event-wise precision or distinctiveness scores.  This enables us to quantify how precisely a participant tended to recall subtle within-event details, as well as how specific (distinctive) those details were to individual events from the episode.  Participants' average precision and distinctiveness scores were strongly correlated ($r(15) = 0.90, p < 0.001$).  This indicates that participants who tended to precisely recount low-level details of episode events also tended to do so in an event-specific way (e.g., as opposed to detailing recurring themes that were present in most or all episode events; this behavior would have resulted in high precision but low distinctiveness).  We found that, across participants, higher precision scores were positively correlated with the numbers of both hand-annotated scenes ($r(15) = 0.60, p = 0.010$) and model-estimated events ($r(15) = 0.90, p < 0.001$) that participants recalled.  Participants' average distinctiveness scores were also marginally correlated with both the hand-annotated ($r(15) = 0.45, p = 0.068$) and model-estimated ($r(15) = 0.71, p = 0.001$) numbers of recalled events.

@@

The neural analyses we carried out (Fig.~\ref{fig:brainz}) also leveraged our geometric framework for characterizing the shapes of the episode and participants' recountings.  We identified one network of regions whose responses tracked with temporal correlations in the conceptual content of the episode (as quantified by topic models applied to a set of annotations about the episode).  This network included orbitofrontal cortex, ventromedial prefrontal cortex, and striatum, among others.  As reviewed by Ranganath and Ritchey (2012)~\cite{RangRitc12}, several of these regions are members of the ``anterior temporal system,'' which has been implicated in assessing and processing the familiarity of ongoing experiences, emotions, social cognition, and reward.  A second network we identified tracked with temporal correlations in the idiosyncratic conceptual content of participants' subsequent recountings of the episode.  This network included occipital cortex, extrastriate cortex, fusiform gyrus, and the precuneus.  Several of these regions are members of the ``posterior medial system''~\citep{RangRitc12}, which has been implicated in matching incoming cues about the current situation to internally maintained ``situation models'' that specify the parameters and expectations inherent to the current situation~\citep{ZackEtal07, ZwaaRadv98}.  Taken together, our results support the notion that these two (partially overlapping) networks work in coordination to make sense of our ongoing experiences, distort them in a way that links them with our prior knowledge and experiences, and encodes those distorted representations into memory for our later use.  Our work also provides a potential framework for modeling and elucidating ``memory schemas''-- i.e., cognitive abstractions that may be applied to multiple related experiences~\citep{GilbMarl17, BaldEtal18}.  For example, the event-level geometric scaffolding of an experience (e.g., Fig.~\ref{fig:trajectory}A) might reflect its underlying schema, and experiences that share similar schemas might have similar shapes.  This could also help explain how brain structures including the ventromedial prefrontal cortex~\citep{GilbMarl17} (Fig.~\ref{fig:brainz}) might acquire or apply schema knowledge across different experiences (i.e., by learning patterns in the schema's shape).

@@

Perhaps the simplest and most widely used measure of memory performance is ``accuracy''---i.e., the proportion of studied (experienced) items (in this case, episode events) that the participant later remembered.  Chen et al. (2017)~\cite{ChenEtal17} used this method to rate each participant's memory quality by computing the proportion of (50 manually identified) scenes mentioned in their recall.  We found a strong across-participants correlation between these independent ratings and the proportion of 30 HMM-identified episode events matched to participants' recalls (Pearson's $r(15) = 0.71, p = 0.002$).  We further considered a number of more nuanced memory performance measures that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.

>>>>

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both Chen et al. (2017)~\cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046,~95\%~CI~[0.25, 0.76]$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013,~95\%~CI~[0.31, 0.80]$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity~\citep{MannKaha12}.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.004,~95\%~CI~[0.31, 0.85]$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015,~95\%~CI~[0.10, 0.83]$) numbers of recalled events.

@@

In addition to examining how precisely and distinctively participants recalled individual events, one may also use these metrics to summarize each participant's performance by averaging across a participant's event-wise precision or distinctiveness scores.  This enables us to quantify how precisely a participant tended to recall subtle within-event details, as well as how specific (distinctive) those details were to individual events from the episode.  Participants' average precision and distinctiveness scores were strongly correlated ($r(15) = 0.90,~p < 0.001,~95\%~CI~[0.66, 0.96]$).  This indicates that participants who tended to precisely recount low-level details of episode events also tended to do so in an event-specific way (e.g., as opposed to detailing recurring themes that were present in most or all episode events; this behavior would have resulted in high precision but low distinctiveness).  We found that, across participants, higher precision scores were positively correlated with the numbers of both hand-annotated scenes ($r(15) = 0.60,~p = 0.010,~95\%~CI~[0.02, 0.83]$) and model-estimated events ($r(15) = 0.90,~p < 0.001,~95\%~CI~[0.54, 0.96]$) that participants recalled.  Participants' average distinctiveness scores were also marginally correlated with both the hand-annotated ($r(15) = 0.45,~p = 0.068,~95\%~CI~[-0.21, 0.79]$) and model-estimated ($r(15) = 0.71,~p = 0.001,~95\%~CI~[-0.07, 0.90]$) numbers of recalled events.

@@

The neural analyses we carried out (Fig.~\ref{fig:brainz}) also leveraged our geometric framework for characterizing the shapes of the episode and participants' recountings.  We identified one network of regions whose responses tracked with temporal correlations in the conceptual content of the episode (as quantified by topic models applied to a set of annotations about the episode).  This network included orbitofrontal cortex, ventromedial prefrontal cortex, and striatum, among others.  As reviewed by Ranganath and Ritchey (2012)~\cite{RangRitc12}, several of these regions are members of the ``anterior temporal system,'' which has been implicated in assessing and processing the familiarity of ongoing experiences, emotions, social cognition, and reward.  A second network we identified tracked with temporal correlations in the idiosyncratic conceptual content of participants' subsequent recountings of the episode.  This network included occipital cortex, extrastriate cortex, fusiform gyrus, and the precuneus.  Several of these regions are members of the ``posterior medial system''~\citep{RangRitc12}, which has been implicated in matching incoming cues about the current situation to internally maintained ``situation models'' that specify the parameters and expectations inherent to the current situation~\citep{ZackEtal07, ZwaaRadv98}.  Taken together, our results support the notion that these two (partially overlapping) networks work in coordination to make sense of our ongoing experiences, distort them in a way that links them with our prior knowledge and experiences, and encodes those distorted representations into memory for our later use.  Our work also provides a potential framework for modeling and elucidating ``memory schemas''---i.e., cognitive abstractions that may be applied to multiple related experiences~\citep{GilbMarl17, BaldEtal18}.  For example, the event-level geometric scaffolding of an experience (e.g., Fig.~\ref{fig:trajectory}A) might reflect its underlying schema, and experiences that share similar schemas might have similar shapes.  This could also help explain how brain structures including the ventromedial prefrontal cortex~\citep{GilbMarl17} (Fig.~\ref{fig:brainz}) might acquire or apply schema knowledge across different experiences (i.e., by learning patterns in the schema's shape).

@@

Perhaps the simplest and most widely used measure of memory performance is ``accuracy''---i.e., the proportion of studied (experienced) items (in this case, episode events) that the participant later remembered.  Chen et al.~(2017)~\cite{ChenEtal17} used this method to rate each participant's memory quality by computing the proportion of (50 manually identified) scenes mentioned in their recall.  We found a strong across-participants correlation between these independent ratings and the proportion of 30 HMM-identified episode events matched to participants' recalls (Pearson's $r(15) = 0.71, p = 0.002,~95\%~CI~[0.39, 0.88]$).  We further considered a number of more nuanced memory performance measures that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.
